A test for diversifying selection for a trait from within and between species variations while integrating for genomic diversity

### 2023-07-13 manuscript/main.tex
Updated manuscript

@@

\title{A test of diversifying selection for a trait from within and between species variations}
\rhead{\scshape Test of diversifying selection}

@@

% Abstract (≤ 250 words)
%TC:endignore
\begin{abstract}
    To determine whether a trait is neutrally evolving or under selection, methods have been developed using either within or between species variation.
    However, using within species variation, methods do not integrate the changes over long evolutionary time.
    Conversely, using between species variation, current methods usually discards the variance in traits between individuals, leaving out valuable information.
    The main goal of this study is to define a neutrality index for a quantitative trait, by combining within- and between-species variation.
    This neutrality index this does not require estimation of population size nor of time of speciation for normalization since it also leverages sequence polymorphism and divergence.
    Our index can be used to seek deviation from the null model of neutral evolution, and is greater than one for a trait under diversifying selection.
    As an example, we show that brain mass at the mammalian scale is under diversifying selection.
    Finally, we show that our test is not sensitive to the assumption that population sizes, mutation rates and generation time are constant across the phylogeny, and automatically adjust for it.
\end{abstract}

@@

% Research Article (7500 words)
\section{Introduction}\label{sec:introduction}

Determining whether a particular trait is neutrally evolving or under a particular regime of selection has been a long-standing goal in evolutionary biology.
Fundamentally, distinguishing neutral evolution from selection requires determining which mode of selection is supported by the observed variation of traits and sequences.
Trait variation can be observed at different scales, across different development stages at the individual level, across different individuals and populations at the species level, and finally across different species at the phylogenetic level.
All these scales require different assumptions and methodologies, and the endeavor to determine the mode of selection for a given trait has thus incorporated theories, methods, and developments across various fields of evolutionary biology such as quantitative-genetics, population-genetics, phylogenetics and comparative genomics\cite{lynch_genetics_1998, walsh_evolution_2018}.

At the population level across individuals, Genome Wide Association Studies (GWAS) in humans have revealed that traits are mostly polygenic (many loci associated to a given trait), pleiotropic (many traits associated to a given loci), additive and mainly under stabilizing selection\cite{simons_population_2018, sella_thinking_2019}.
At the species level, by contrasting both trait and genetic differentiation across populations, Q$_\text{ST}$--F$_\text{ST}$ methods have allowed to quantify selection acting on a trait\cite{martin_multivariate_2008, leinonen_qst_2013}
For example, studies have shown that stabilizing selection is largely dominant in the evolution of expression of genes in different populations\cite{whitehead_neutral_2006, gilad_natural_2006}.
However, dominant mode of selection acting on gene expression is still controversial and neutral evolution is still debated\cite{signor_evolution_2018, price_detecting_2022}.
Theoretically, change in traits accumulate linearly with time from divergence for a pair of species, and proportionally to the trait variance\cite{lande_genetic_1980, turelli_heritable_1984}.
For example, gene expression accumulate linearly with time from divergence, and accumulates faster for trait with large variance\cite{khaitovich_neutral_2004}.
Such theoretical connection between trait variance and change in mean value can be used to test for a selection of trait using for a pair of species\cite{walsh_evolution_2018}.
However, to our knowledge, a neutral theoretical expectation for trait changes across several species related by their phylogenetic tree, as a function of within species trait variation has not yet been formulated and tested.

% Phylogenetics (mean trait evolution)
Contrarily, by accounting for the underlying relationships between species, the mode of selection for a quantitative trait can also be tested at the phylogenetic scale\cite{felsenstein_phylogenies_1985}.
Indeed, if the trait is neutrally evolving, the change in main trait value along a given branch of the tree should be Normally distributed around the ancestral state (at the beginning of branch), with a standard deviation proportional to the squared root of the divergence time\cite{hansen_translating_1996}.
As a result, the mean trait value along the whole phylogeny can be modeled as a Brownian process branching at every node of the tree, allowing theoretically to test for neutrality\cite{hansen_translating_1996, harmon_phylogenetic_2018}.
A deviation from the Brownian process, typically an Ornstein-Uhlenbeck process, is interpreted as a signature of stabilizing selection\cite{catalan_drift_2019}.
Alternatively, a rapid shift in mean trait value along a branch is interpreted as a signature of directional selection.
However, studies have shown that such comparative approaches are subject to different biases\cite{harmon_phylogenetic_2018}.
First, and unfortunately, a better fit for a Brownian process at the phylogenetic scale is not necessarily a proof of the neutral model.
A trait under stabilizing selection for which the optimal trait value is also evolving as a Brownian process will not deviate from a Brownian process, and thus be wrongly classified as neutral\cite{hansen_translating_1996}.
Secondly and contrarily, even under a neutral trait, the Ornstein-Uhlenbeck process might sometimes be statistically preferred over a Brownian process due to sampling artifacts\cite{silvestro_measurement_2015, cooper_cautionary_2016, price_detecting_2022}.
In such comparative framework, the variation within species is usually discarded, where the mean trait value is taken either as the trait for a single individual in the species or as the average trait across several individuals.
Altogether, methods at the phylogenetic scale easily mis-classify selection and also discards the variance in traits between individuals, leaving out valuable information.

% At the frontier Phylogenetics/Population-genetics
At the frontier between phylogenetics and population genetics, comparative methods at the phylogenetic scale have acknowledged the importance to model within species variation on top of changes in mean trait value\cite{fay_evaluating_2008, kostikova_bridging_2016, gaboriau_multiplatform_2020}.
However, within species variation is used to describe technical errors, or the ratio of between to within variation is estimated for many traits, and is finally compared to the average over all traits to seek deviation from this average\cite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015}.
Determining the evolutionary regime of a quantitative trait through articulation of the variance between and within species is the goal is this study, while setting threshold for neutral, stabilizing and diversifying selection.
In this study, we propose a neutrality index for a quantitative trait articulating trait and sequence variation within and between species.
Importantly, our neutrality index also leverages sequence divergence and polymorphism in order to normalize trait variation at both scales, such that is does not requires estimate of population size (within species) nor of time of speciation (between species).
From a population-genetic perspective, our study can be seen as the generalization of Q$_\text{ST}$--F$_\text{ST}$ methods at the phylogenetic scale to account relationships between species.
From a phylogenetic perspective, our study can be seen as a extension of the EVE model\cite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015} where we set a clear threshold for neutral evolution by leveraging species' sequence polymorphism and divergence.
% Basically, we are testing whether the variance of means is greater than the man of variances

\section{Materials and Methods}\label{sec:materials-and-methods}
\subsection{Neutrality index for a quantitative trait}\label{subsec:neutrality-index-for-a-quantitative-trait}
\subsubsection{Mutation-drift equilibrium for a trait}

@@

\subsubsection{Phylogenetic evolution of a trait}

Along a lineage, we denote $\MeanTrait$ as the mean value of the trait over the individuals in the species.
If the trait is neutral and encoded by many loci, $\MeanTrait$ evolves as Brownian process\cite{hansen_translating_1996}.
After the species has evolved for $\Time$ generations, the variance of $\MeanTrait$, called $\VarPhy$, is given by \textcite{hansen_translating_1996} as:
\begin{align}
    \VarPhy & = \frac{\VarGenetic}{\Ne} \Multiply \Time \\
    & = 4 \Time \Multiply \MutationRate \Multiply \RateMut \label{eq:covar},\text{ from eq.~\ref{eq:var-genetic}},
\end{align}

Moreover, for any neutral genomic region of interest, out of all neutral mutations that originated, some will eventually reach fixation due to genetic drift, resulting in a substitution at the species level.
For a neutral mutation, its probability of fixation ($\pfix$) is $1/2\Ne$, as shown in \textcite{kimura_probability_1962}.
As reviewed in \textcite{mccandlish_modeling_2014}, we can thus derive the substitution rate per generation, denoted $\SubRate$ as the number of mutations per generation ($2\Ne \Multiply \MutationRate$) multiplied by the probability of fixation for each of these newly arisen mutations $\pfix$, giving:
\begin{align}
    \SubRate & = 2 \Ne \Multiply \MutationRate \Multiply \pfix, \\
    & = 2 \Ne  \Multiply \MutationRate  \Multiply \dfrac{1}{2\Ne}, \\
    & = \MutationRate. \label{eq:substitution-rate}
\end{align}
That is, if mutations are neutral, the rate with which a new neutral allele are fixed in a genomic region of interest equals the rate with which new mutations arise per generation for the same genomic region\cite{kimura_evolutionary_1968}.

@@

\subsubsection{Neutrality index}

@@

\subsection{Estimate}

\subsubsection{Maximum likelihood estimate}
At the phylogenetic scale, for $\NbrTaxa$ taxa in the tree, $\DistanceMatrix$ ($\NbrTaxa \times \NbrTaxa$) is the distance matrix computed from the branch lengths ($d$) and the topology of the phylogenetic tree.
Along the diagonal $\Distance_{\Spi,\Spi}$ are the total distances from the root of the tree to each taxon ($\Spi$).
The distance matrix is symmetric and the off-diagonal elements ($\Distance_{\Spi,\Spj}$) are total distance between the root and the common ancestor of taxa $\Spi$ and $\Spj$ (which can be root, in which case the distance is 0).
Then, the root state for the character, called $\RootTrait$, can be estimated from the $\NbrTaxa \times 1$ vector of mean trait values for tip species in the tree, called $\VecTrait$, as a maximum likelihood estimate\cite{omeara_testing_2006}, given as:
\begin{gather}
    \RootTrait = \left( \VecOne\tr \MultiplyMatrix \DistanceMatrix\inv \MultiplyMatrix \VecOne \right)\inv \Multiply \left( \VecOne\tr \MultiplyMatrix \DistanceMatrix\inv \MultiplyMatrix \VecTrait \right), \label{eq:estimated-root-trait}
\end{gather}
where $\VecOne$ is an $\NbrTaxa \times 1$ column vector of ones.

Finally, the normalized mutational variance at the phylogenetic scale ($\EstRatePhy$) is estimated\cite{omeara_testing_2006} as:
\begin{gather}
    \EstRatePhy = \frac{1}{4}\frac{\left( \VecTrait -  \RootTrait \Multiply \VecOne \right)\tr \MultiplyMatrix \DistanceMatrix\inv \MultiplyMatrix \left( \VecTrait -  \RootTrait \Multiply \VecOne  \right)}{\NbrTaxa - 1}. \label{eq:estimated-rate-phy}
\end{gather}

@@

\subsubsection{Bayesian estimate}

Even though the neutrality index can be applied for has many traits as available, each trait is so far assumed to be independent.
Assuming covariation between traits is also an extension that can be implemented within the multidimensional Brownian framework~\cite{huelsenbeck_detecting_2003, lartillot_phylogenetic_2011, lartillot_joint_2012, latrille_inferring_2021}.
Here we generalize to $\Ntrait$ traits evolving along the phylogeny and are correlated between them using the \textit{BayesCode} software~\cite{latrille_inferring_2021} .
Their variation along the phylogeny is modelled as an $\Ntrait$-dimensional Brownian process $\Brownian$ ($1 \times \Ntrait$) starting a the root and branching along the tree topology.
The rate of change of the Brownian process is determined by the positive semi-definite and symmetric covariance matrix between traits $\CovarianceMatrix$ ($\Ntrait \times \Ntrait$).
The off-diagonal elements of $\CovarianceMatrix$ are the covariance between traits, and the diagonal elements are the variance of each trait, thus corresponding to $\NI$.
We defined the {prior} on the covariance matrix as an inverse Wishart distribution, the {posterior} on $\CovarianceMatrix$, conditional on a particular realization of $\brownian$ is also an invert Wishart distribution.
The posterior distribution of $\EstNI$ thus allow to test for deviation from neutrality, for example by computing $\proba{\NI > 1 }$ to test for evidence of diversifying selection and $\proba{\NI < 1 }$ to test for evidence of stabilizing selection.

\subsection{Simulation}\label{subsec:simulations}

We will assume that the trait genotypic value is encoded by $\NbrLoci$ loci, with each locus contributing to the genotypic value, these effect are Normally distributed with standard deviation $a$ (Fig.~\ref{fig:simulator}).
We will follow a Wright-Fisher model with mutation, selection and drift to formalise the evolutionary dynamics of the system.
Assuming a diploid panmictic population of size $\Ne$ and with non-overlapping generations, we can simulate explicitly each generation along a phylogenetic tree.
Genetic drift is modelled by the multinomial resampling of the currently segregating alleles.
The number of mutations is drawn from a Poisson distribution with mean $2 \Multiply \MutationRate \Multiply \NbrLoci $.
The changes in the mutation rate per generation ($\MutationRate$) and $\Ne$ along the lineages is modelled by a geometric Brownian process.
Additionally, the instant value of log-$\Ne$ is modelled as a sum of a geometric Brownian process and an Ornstein-Uhlenbeck process.
The geometric Brownian motion accounts for long-term fluctuations, while the Ornstein-Uhlenbeck introduces short-term fluctuations.
The simulations will start from an initial sequence at equilibrium at the root of the tree and, at each node, the process will be split until it finally reaches the leaves of the tree.
From a speciation process perspective, this a equivalent to an allopatric speciation over one generation.
Selection is modelled as a one dimensional Fisher's geometric landscape, with the fitness of an individual being a monotonously decreasing function of the distance between the individual and optimal phenotype~\cite{tenaillon_utility_2014,blanquart_epistasis_2016}.
The exact functional phenotype-fitness map depends on $2$ external parameters controlling for strength of selection and optimum phenotype.
Mutations are seen has displacement of the phenotype in the multidimensional space.
Beneficial mutations moves the phenotype closer to the optimum, while deleterious mutations moves it further away.
Stabilizing selection is implemented by means of a fixed optimum phenotype.
Diversifying selection is implemented by means of a moving optimum phenotype, changing along the phylogeny as a geometric Brownian process.
Genetic diversity is measured as the heterozygosity of the neutral markers, and the genetic variance of the trait is measured as the variance of the trait genotypic value.
Genetic divergence is measured as the number of substitutions per site along branches of the phylogeny.
Heritability is estimated from the slope of the regression of offspring trait values on parental trait values\cite{lynch_genetics_1998}, using the information from the last two generations of the simulation.

\begin{figure*}[!ht]
    \centering
    \includegraphics[width=\textwidth, page=1] {artworks/fig-simulator}
    \caption{
        Wright-Fisher simulations with mutation, selection and drift.
    }
    \label{fig:simulator}
\end{figure*}

\subsection{Empirical dataset}

@@

\section{Results}

To determine whether a trait is neutrally evolving or under selection, we reviewed the theoretical equations of trait evolution at different scales.
The underlying mutational input for a neutral trait is formally related to both within and between species variation of the trait.
More specifically, for a heritable trait under a neutral regime, the genetic variance of the trait ($\VarGenetic$) normalized by genetic diversity of any neutral genomic region ($\pi$) is a proxy for the mutational variance ($\RatePop$).
At the phylogenetic level, along a specific lineage, the mean value for the trait evolves as a Brownian process, and the variance in this mean value after a time $\Time$ is called $\VarPhy$.
This variance can also be normalized by the genetic distance of any neutral genomic region ($d$), allowing us to obtain another proxy for the mutational variance ($\RatePhy$).
Altogether, we define the neutrality index as $\NI = \RatePhy/\RatePop$, which equals to $1$ for a neutral trait, suggesting that traits for which this relationship is not verified are putatively under a selective regime.
Under a depleted variation between species, meaning $\NI < 1$, the trait is putatively under stabilizing, forcing the mean value to be homogeneous between species.
In contrast, under an inflated variation between species, meaning $\NI > 1$, the trait is putatively under diversifying selection, or a moving optimum such that the mean value is heterogeneous in the different species.

>>>>

% -- Defining colors:
\definecolor{backcolour}{rgb}{0.95,0.95,0.92}% Definig a custom style:
\lstdefinestyle{mystyle}{
    backgroundcolor=\color{backcolour},
    basicstyle=\ttfamily\scriptsize\bfseries,
    breakatwhitespace=false,
    breaklines=true,
    captionpos=t,
    keepspaces=true,
    showspaces=false,
    showstringspaces=false,
    showtabs=false,
    tabsize=2
}% -- Setting up the custom style:
\lstset{style=mystyle}

@@

\title{A test for diversifying selection for a trait from within and between species variations while integrating for genomic diversity}
\rhead{\scshape Test of diversifying selection}

@@

% Abstract (≤ 250 words)
%TC:endignore
\begin{abstract}
    To determine whether a trait is neutrally evolving or under selection, methods have been developed using either within or between species variation.
    Using within species variation, however, methods do not integrate the changes at the macro-evolutionary scale.
    Conversely, using between species variation, current methods usually discards within species variation, thus not accounting for processes at the micro-evolutionary scale.
    The main goal of this study is to define a neutrality index for a quantitative trait, by combining within- and between-species variation.
    This neutrality index integrates nucleotide polymorphism and divergence for normalizing trait variation.
    As such, it does not require estimation of population size nor of time of speciation for normalization.
    Our index can be used to seek deviation from the null model of neutral evolution, and test for diversifying selection.
    Applied to brain mass and body mass at the mammalian scale, we show that brain mass is under diversifying selection.
    Finally, we show that our test is not sensitive to the assumption that population sizes, mutation rates and generation time are constant across the phylogeny, and automatically adjust for it.
\end{abstract}

@@

% Research Article (7500 words)
\section*{Introduction}\label{sec:introduction}

Determining whether a particular trait is neutrally evolving or under a particular regime of selection has been a long-standing goal in evolutionary biology.
Fundamentally, distinguishing neutral evolution from selection requires determining which mode of selection is supported by the observed variation of traits and sequences.
Trait variation can be observed at different scales, across different development stages at the individual level, across different individuals and populations at the species level, and finally across different species at the phylogenetic level.
All these scales require different assumptions and methodologies, and the endeavor to determine the mode of selection for a given trait has thus incorporated theories, methods, and developments across various fields of evolutionary biology such as quantitative-genetics, population-genetics, phylogenetics and comparative genomics~\cite{lynch_genetics_1998, walsh_evolution_2018}.

First, at the micro-evolutionary scale, leveraging individuals variation within the same species, Genome Wide Association Studies (GWAS) in humans have shown that traits are mostly polygenic (many loci associated to a given trait), pleiotropic (many traits associated to a given loci), additive and mainly under stabilizing selection~\cite{simons_population_2018, sella_thinking_2019}.
At the species level, by contrasting both trait and genetic differentiation across populations, Q$_\text{ST}$--F$_\text{ST}$ methods have allowed to quantify selection acting on a trait~\cite{martin_multivariate_2008, leinonen_qst_2013}.
For example, studies have shown that stabilizing selection is largely dominant in the evolution of expression of genes in different populations~\cite{whitehead_neutral_2006, gilad_natural_2006}.
However, dominant mode of selection acting on gene expression is still controversial and neutral evolution is still debated~\cite{signor_evolution_2018, price_detecting_2022}.
Theoretically, change in traits accumulate linearly with time from divergence for a pair of species, and proportionally to the trait variance~\cite{lande_genetic_1980, turelli_heritable_1984}.
For example, gene expression accumulate linearly with time from divergence, and accumulates faster for trait with large variance~\cite{khaitovich_neutral_2004}.
Such theoretical connection between trait variance and change in mean value can be used to test for a selection of trait using for a pair of species~\cite{walsh_evolution_2018}.
However, to our knowledge, a neutral theoretical expectation for trait changes across several species related by their phylogenetic tree, as a function of within species trait variation has not yet been formulated and tested.

% Phylogenetics (mean trait evolution)
Contrarily, at the macro-evolutionary scale, by accounting for the underlying relationships between species, the mode of selection for a quantitative trait can also be tested at the phylogenetic scale~\cite{felsenstein_phylogenies_1985}.
Indeed, if the trait is neutrally evolving, the change in main trait value along a given branch of the tree should be Normally distributed around the ancestral state (at the beginning of branch), with a standard deviation proportional to the squared root of the divergence time~\cite{hansen_translating_1996}.
As a result, the mean trait value along the whole phylogeny can be modeled as a Brownian process branching at every node of the tree, allowing theoretically to test for neutrality~\cite{hansen_translating_1996, harmon_phylogenetic_2018}.
A deviation from the Brownian process, typically an Ornstein-Uhlenbeck process, is interpreted as a signature of stabilizing selection~\cite{catalan_drift_2019}.
Alternatively, a rapid shift in mean trait value along a branch is interpreted as a signature of directional selection.
However, studies have shown that such comparative approaches are subject to different biases~\cite{harmon_phylogenetic_2018}.
First, and unfortunately, a better fit for a Brownian process at the phylogenetic scale is not necessarily a proof of the neutral model.
A trait under stabilizing selection for which the optimal trait value is also evolving as a Brownian process will not deviate from a Brownian process, and thus be wrongly classified as neutral~\cite{hansen_translating_1996}.
Secondly and contrarily, even under a neutral trait, the Ornstein-Uhlenbeck process might sometimes be statistically preferred over a Brownian process due to sampling artifacts~\cite{silvestro_measurement_2015, cooper_cautionary_2016, price_detecting_2022}.
In such comparative framework, the variation within species is usually discarded, where the mean trait value is taken either as the trait for a single individual in the species or as the average trait across several individuals.
Altogether, methods at the phylogenetic scale easily mis-classify selection and also discards the variance in traits between individuals, leaving out valuable information.

% At the frontier Phylogenetics/Population-genetics
At the frontier between phylogenetics and population genetics, comparative methods at the phylogenetic scale have acknowledged the importance to model within species variation on top of changes in mean trait value~\cite{fay_evaluating_2008, kostikova_bridging_2016, gaboriau_multiplatform_2020}.
However, within species variation is used to describe technical errors, or the ratio of between to within variation is estimated for many traits, and is finally compared to the average over all traits to seek deviation from this average~\cite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015}.
Determining the evolutionary regime of a quantitative trait through articulation of the variance between and within species is the goal is this study, while setting threshold for neutral, stabilizing and diversifying selection.
In this study, we propose a neutrality index for a quantitative trait articulating trait and sequence variation within and between species.
Importantly, our neutrality index also leverages sequence divergence and polymorphism in order to normalize trait variation at both scales, such that is does not requires estimate of population size (within species) nor of time of speciation (between species).
From a population-geneticist perspective, our study can be seen as the generalization at the macro scale of Q$_\text{ST}$--F$_\text{ST}$ methods to account for phylogenetic relationships between species.
From a phylogeneticist perspective, our study can be seen as a extension of the EVE model~\cite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015} where we set a clear threshold for neutral evolution by leveraging species' sequence polymorphism and divergence.
% Basically, we are testing whether the variance of means is greater than the man of variances

\section*{Materials and Methods}\label{sec:materials-and-methods}
\subsection*{Neutrality index for a quantitative trait}\label{subsec:neutrality-index-for-a-quantitative-trait}
\subsubsection*{Mutation-drift equilibrium for a trait}

@@

\subsubsection*{Phylogenetic evolution of a trait}

Along a lineage, we denote $\MeanTrait$ as the mean value of the trait over the individuals in the species.
If the trait is neutral and encoded by many loci, $\MeanTrait$ evolves as Brownian process~\cite{hansen_translating_1996}.
After the species has evolved for $\Time$ generations, the variance of $\MeanTrait$, called $\VarPhy$, is given by \textcite{hansen_translating_1996} as:
\begin{align}
    \VarPhy & = \frac{\VarGenetic}{\Ne} \Multiply \Time \\
    & = 4 \Time \Multiply \MutationRate \Multiply \RateMut \label{eq:covar},\text{ from eq.~\ref{eq:var-genetic}},
\end{align}

Moreover, for any neutral genomic region of interest, out of all neutral mutations that originated, some will eventually reach fixation due to genetic drift, resulting in a substitution at the species level.
For a neutral mutation, its probability of fixation ($\pfix$) is $1/2\Ne$, as shown in \textcite{kimura_probability_1962}.
As reviewed in \textcite{mccandlish_modeling_2014}, we can thus derive the substitution rate per generation, denoted $\SubRate$ as the number of mutations per generation ($2\Ne \Multiply \MutationRate$) multiplied by the probability of fixation for each of these newly arisen mutations $\pfix$, giving:
\begin{align}
    \SubRate & = 2 \Ne \Multiply \MutationRate \Multiply \pfix, \\
    & = 2 \Ne  \Multiply \MutationRate  \Multiply \dfrac{1}{2\Ne}, \\
    & = \MutationRate. \label{eq:substitution-rate}
\end{align}
That is, if mutations are neutral, the rate with which a new neutral allele are fixed in a genomic region of interest equals the rate with which new mutations arise per generation for the same genomic region~\cite{kimura_evolutionary_1968}.

@@

\subsubsection*{Neutrality index}

@@

\subsection*{Estimate}\label{subsec:estimate}

\subsubsection*{Maximum likelihood estimate}
At the phylogenetic scale, for $\NbrTaxa$ taxa in the tree, $\DistanceMatrix$ ($\NbrTaxa \times \NbrTaxa$) is the distance matrix computed from the branch lengths ($d$) and the topology of the phylogenetic tree.
Along the diagonal $\Distance_{\Spi,\Spi}$ are the total distances from the root of the tree to each taxon ($\Spi$).
The distance matrix is symmetric and the off-diagonal elements ($\Distance_{\Spi,\Spj}$) are total distance between the root and the common ancestor of taxa $\Spi$ and $\Spj$ (which can be root, in which case the distance is 0).
Then, the root state for the character, called $\RootTrait$, can be estimated from the $\NbrTaxa \times 1$ vector of mean trait values for tip species in the tree, called $\VecTrait$, as a maximum likelihood estimate~\cite{omeara_testing_2006}, given as:
\begin{gather}
    \RootTrait = \left( \VecOne\tr \MultiplyMatrix \DistanceMatrix\inv \MultiplyMatrix \VecOne \right)\inv \Multiply \left( \VecOne\tr \MultiplyMatrix \DistanceMatrix\inv \MultiplyMatrix \VecTrait \right), \label{eq:estimated-root-trait}
\end{gather}
where $\VecOne$ is an $\NbrTaxa \times 1$ column vector of ones.

Finally, the normalized mutational variance at the phylogenetic scale ($\EstRatePhy$) is estimated~\cite{omeara_testing_2006} as:
\begin{gather}
    \EstRatePhy = \frac{1}{4}\frac{\left( \VecTrait -  \RootTrait \Multiply \VecOne \right)\tr \MultiplyMatrix \DistanceMatrix\inv \MultiplyMatrix \left( \VecTrait -  \RootTrait \Multiply \VecOne  \right)}{\NbrTaxa - 1}. \label{eq:estimated-rate-phy}
\end{gather}

@@

\subsubsection*{Bayesian estimate}

Even though the neutrality index can be applied for has many traits as available, each trait is so far assumed to be independent.
Assuming covariation between traits is also an extension that can be implemented within the multidimensional Brownian framework~\cite{huelsenbeck_detecting_2003, lartillot_phylogenetic_2011, lartillot_joint_2012, latrille_inferring_2021}.
Here we generalize to $\Ntrait$ traits evolving along the phylogeny and are correlated between them using the \textit{BayesCode} software~\cite{latrille_inferring_2021}.
Their variation along the phylogeny is modelled as an $\Ntrait$-dimensional Brownian process $\Brownian$ ($1 \times \Ntrait$) starting a the root and branching along the tree topology (see section~S\ref{subsec:multivariate-brownian-process}).
The rate of change of the Brownian process is determined by the positive semi-definite and symmetric covariance matrix between traits $\CovarianceMatrix$ ($\Ntrait \times \Ntrait$).
The off-diagonal elements of $\CovarianceMatrix$ are the covariance between traits, and the diagonal elements are the variance of each trait, thus corresponding to $\NI$.
We defined the {prior} on the covariance matrix as an inverse Wishart distribution, the {posterior} on $\CovarianceMatrix$, conditional on a particular realization of $\brownian$ is also an invert Wishart distribution (see section~S\ref{subsec:sampling-the-covariance-matrix}).
The posterior distribution of $\EstNI$ thus allow to test for deviation from neutrality, for example by computing $\proba [\NI > 1 ]$ to test for evidence of diversifying selection and $\proba [\NI < 1 ]$ to test for evidence of stabilizing selection.

\subsection*{Simulation}\label{subsec:simulations}

We will assume that the trait genotypic value is encoded by $\NbrLoci$ loci, with each locus contributing to the genotypic value, these effect are Normally distributed with standard deviation $a$ (Fig.~\ref{fig:simulator} and section~S\ref{subsec:genotype-phenotype-map}).
We will follow a Wright-Fisher model with mutation, selection and drift to formalise the evolutionary dynamics of the system.
Assuming a diploid panmictic population of size $\Ne$ and with non-overlapping generations, we can simulate explicitly each generation along a phylogenetic tree.
Genetic drift is modelled by the multinomial resampling of the currently segregating alleles.
The number of mutations is drawn from a Poisson distribution with mean $2 \Multiply \MutationRate \Multiply \NbrLoci $.
The changes in the mutation rate per generation ($\MutationRate$) and $\Ne$ along the lineages is modelled by a geometric Brownian process.
Additionally, the instant value of log-$\Ne$ is modelled as a sum of a geometric Brownian process and an Ornstein-Uhlenbeck process.
The geometric Brownian motion accounts for long-term fluctuations, while the Ornstein-Uhlenbeck introduces short-term fluctuations.
The simulations will start from an initial sequence at equilibrium at the root of the tree and, at each node, the process will be split until it finally reaches the leaves of the tree.
From a speciation process perspective, this a equivalent to an allopatric speciation over one generation.
Selection is modelled as a one dimensional Fisher's geometric landscape, with the fitness of an individual being a monotonously decreasing function of the distance between the individual and optimal phenotype~\cite{tenaillon_utility_2014,blanquart_epistasis_2016}.
The exact functional phenotype-fitness map depends on $2$ external parameters controlling for strength of selection and optimum phenotype (see section~S\ref{subsec:phenotype-fitness-map}).
Mutations are seen has displacement of the phenotype in the multidimensional space.
Beneficial mutations moves the phenotype closer to the optimum, while deleterious mutations moves it further away.
Stabilizing selection is implemented by means of a fixed optimum phenotype.
Diversifying selection is implemented by means of a moving optimum phenotype, changing along the phylogeny as a geometric Brownian process.
Genetic diversity is measured as the heterozygosity of the neutral markers, and the genetic variance of the trait is measured as the variance of the trait genotypic value.
Genetic divergence is measured as the number of substitutions per site along branches of the phylogeny.
Heritability is estimated from the slope of the regression of offspring trait values on parental trait values~\cite{lynch_genetics_1998}, using the information from the last two generations of the simulation.

\begin{figure*}[!ht]
    \centering
    \includegraphics[width=\textwidth, page=1] {artworks/fig-simulator}
    \caption{
        Wright-Fisher simulations with mutation, selection and drift.
        Panel A: the trait genotypic value is encoded by $\NbrLoci$ loci, with each locus contributing additively to the genotypic value.
        The trait genotypic value is then transformed into a phenotypic value by adding a Gaussian noise with standard deviation $\VarEnv$.
        Parents are selected for reproduction according to their phenotypic value, with a probability proportional to their fitness.
        Mutations are drawn from a Poisson distribution, with each loci having a probability $\MutationRate$ to mutate.
        Drift is modelled by the resampling of parents in a finite population.
        Panel B: example of a trait evolving along a phylogeny, with the mean phenotype (black line) and the variance of the trait genotypic value (grey area).
    }
    \label{fig:simulator}
\end{figure*}

\subsection*{Empirical dataset}\label{subsec:empirical-dataset}

@@

\section*{Results}\label{sec:results}

To determine whether a trait is neutrally evolving or under selection, we reviewed the theoretical equations of trait evolution at different scales.
The underlying mutational input for a neutral trait is formally related to both within and between species variation of the trait.
More specifically, for a heritable trait under a neutral regime, the genetic variance of the trait ($\VarGenetic$) normalized by genetic diversity of any neutral genomic region ($\pi$), denoted $\RatePop$, is a proxy for the mutational variance ($\RateMut$).
At the phylogenetic level, along a specific lineage, the mean value for the trait evolves as a Brownian process, and the variance in this mean value after a time $\Time$ is called $\VarPhy$.
This variance can also be normalized by the genetic distance of any neutral genomic region ($d$), denoted $\RatePhy$, allowing us to obtain another proxy for the mutational variance ($\RateMut$).
Altogether, we define the neutrality index as $\NI = \RatePhy/\RatePop$, which equals to $1$ for a neutral trait, suggesting that traits for which this relationship is not verified are putatively under a selective regime.
Under a depleted variation between species, meaning $\NI < 1$, the trait is putatively under stabilizing, forcing the mean value to be homogeneous between species.
In contrast, under an inflated variation between species, meaning $\NI > 1$, the trait is putatively under diversifying selection, or a moving optimum such that the mean value is heterogeneous in the different species.

\begin{table*}[t!]
    \centering
    \begin{adjustbox}{width = 1.0\textwidth}
    \begin{tabular}{||l|l||c|c||c|c||}
            \hline
            Assumption & Consequences & $\EstRatePop$ & $\EstRatePhy$ & Test $\NI > 1$  & Test $\NI < 1$ \\ \hline \hline
            Trait encoded by few loci & Between species trait variation is under-estimated  & -- & Under-estimated & Conservative & Invalid  \\ \hline
            Sexual dimorphism &  Within species trait variation is over-estimated  & Over-estimated & -- & Conservative & Invalid  \\ \hline
            Inbreeding & Nucleotide diversity ($\pi$) is under-estimated  & Over-estimated & -- & Conservative & Invalid  \\ \hline
            Markers for polymorphism are negatively selected & Nucleotide diversity ($\pi$) is under-estimated  & Over-estimated & -- & Conservative & Invalid  \\ \hline
            Markers for divergence are positively selected & Nucleotide divergence ($d$) is over-estimated & -- & Under-estimated & Conservative & Invalid  \\ \hline
            Markers for polymorphism are positively selected & Nucleotide diversity ($\pi$) is over-estimated  & Under-estimated & -- & Invalid & Conservative  \\ \hline
            Markers for divergence are negatively selected & Nucleotide divergence ($d$) is under-estimated & -- & Over-estimated & Invalid & Conservative  \\ \hline
    \end{tabular}
    \end{adjustbox}
    \caption{Assumptions and their consequences on the estimation of within species variation ($\EstRatePop$), between species variation ($\EstRatePhy$), and on the neutrality index $\NI = \EstRatePhy/\EstRatePop$.
    The two last columns indicate whether the test for diversifying selection ($\NI > 1$) and for stabilizing selection $\NI < 1$ are conservative or invalid due to violated assumptions.
    }
    \label{table:assumptions}
\end{table*}
